---
title: "Pruning & Rooting Phylogenetic Trees with ggTree"
author: "Daphne Garcia"
date: "2025-03-31"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      fig.align = "center",
                      # Always relevant to the document directory 
                      # send any figure output to this folder 
                      fig.path = "../figures/04_PhylogeneticTree/",
                      warning = FALSE) 
```


## Timing, seed, load packages

```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()

# Any number can be chosen 
set.seed(238428)

pacman::p_load(tidyverse, phyloseq, ggtree, phytools, tidytree, phylobase,
               install = FALSE)
```


# 1. Load Data files 
```{r load-data}
# Preprocessed phyloseq object 
load("data/03_PreProcessing/raw_preprocessed_physeq.RData")

# Inspect the phyloseq object 
raw_preprocessed_physeq

# Load in the tree! 
unrooted_tree <- read.tree("data/04_PhylogeneticTree/ASVs_unrooted.tree")

# Take a look at the tree
unrooted_tree
str(unrooted_tree)
```


# 2. Merge Unrooted Tree & `raw_preprocessed_physeq`

```{r merge-physeq}
# Intuition check 
stopifnot(ntaxa(raw_preprocessed_physeq) == ntaxa(unrooted_tree))
# No news is good news! 

# Merge the tree with the phyloseq object 
unrooted_physeq <- 
  merge_phyloseq(raw_preprocessed_physeq, unrooted_tree)

# Let's take a look! 
unrooted_physeq
```

# 3. Visualize and inspect tree with [ggtree]

## Kingdom Level Tree

```{r plot-tree-unrooted, fig.width=4.5, fig.height=7}
# Make a basic tree with the domains 
kingdom_tree <- 
  ggtree(unrooted_physeq) + 
  # color tips by kingdom
  geom_tippoint(mapping = aes(color = Kingdom)) + 
  # add a title
  labs(title = "Unrooted Tree") + 
  scale_color_manual(values = c("firebrick3", "cornflowerblue", "grey")) + 
  # Move the legend to the bottom of the tree 
  theme(legend.position = "bottom")

# Look at it 
kingdom_tree

# Check if tree is rooted
is.rooted(unrooted_tree) # should be TRUE
```

# 4. Long Branches? 

INTERPRETATION #1: Are there any suspect branches in the raw, unrooted tree? 
Or does your tree look “clean”? Why or why not?


There are two branches that are much longer than the rest. First, there is one 
node composed of mostly Eukaryota, and one visible NA. Another node has a few 
eukaryotes, and one archaea that seems to be closer, or more related to bacteria.
Hopefully once I get rid of the eukaryotes, the archaea branch will be visibly 
more distant than the bacteria, but currently it seems as if the archaea branch 
is not really distant from bacteria.


## 4a. Evaluate Long Branches

### Kingdom Tree with Node Labels 

Let's put labels on our tree so we have a bit more flexibility

```{r kingdom-node-tree, fig.width=4.5, fig.height=7}
kingdom_node_tree <- 
  kingdom_tree + 
  # Add the node label 
  geom_text(aes(label=node), hjust= -0.5, vjust = -0.3, size = 2)

# Plot the tree
kingdom_node_tree
```
Looks like 23,335 and 23,109 are near the base of the two really long branches


### view specific branches: 23,335 and 23,109
```{r view specific node 23335}
# View a specific clade
# Zoom in on node that looks potentially suspicious
viewClade(kingdom_node_tree + 
          labs(title = "Kingdom Tree: Node 23335"), 
          node = 23335)
```
23335 is a whole branch of eukaryotes!!

```{r view specific node 23109}
# View a specific clade
# Zoom in on node that looks potentially suspicious
viewClade(kingdom_node_tree + 
          labs(title = "Kingdom Tree: Node 23109"), 
          node = 23109)
```

### inspect zoomed out branches of 23335 and 23109

```{r zoom out trees}

#zoom out of 23335
tidytree::ancestor(unrooted_tree, 23335) # The input is the TREE, not phyloseq!!
  # looks like 23089 is MSA of 23335 and 23109!
  # Let's evaluate: 23324 
  viewClade(kingdom_node_tree, 23325) # even more zoomed out
  
  ## 23335 captures all of the eukaryotes in this branch

#zoom out of 23109
tidytree::ancestor(unrooted_tree, 23109) # The input is the TREE, not phyloseq!!
  # looks like 23089 is MSA of 23335 and 23109!
  # Let's evaluate: 23324 
  viewClade(kingdom_node_tree, 23090) # even more zoomed out
  ## 23109 captures all of the eukaryotes in this branch BUT NOT ARCHAEA
  
  
#zoom out of 5251
tidytree::ancestor(unrooted_tree, 5251) # The input is the TREE, not phyloseq!!
  # looks like 23089 is MSA of 23335 and 23109!
  # Let's evaluate: 23324 
  viewClade(kingdom_node_tree, 23091) # even more zoomed out
  ## 5251 captures the one archaea of this branch
  
  # Zoom in on archaea node
viewClade(kingdom_node_tree + 
          labs(title = "Kingdom Tree: Node 5251"), 
          node = 5251)
```


### prepare nodes for 23335 and 23109
```{r tree tibble for node 23335}
# Or if there are two nodes that we can see we can find its MRCA
## create tibble of tree to identify offspring
tree_tib <- as_tibble(unrooted_physeq@phy_tree)

# Inspect
head(tree_tib)

str(tree_tib)

# lets look at the long branch
mrca_node_23335 <- tidytree::MRCA(unrooted_tree, .node1 = 23335) 

# Inspect
mrca_node_23335 

## create dataframe to view ASVs and taxa info and to pull the ASVs
node_23335 <- 
  offspring(tree_tib, mrca_node_23335, tiponly = TRUE) %>% 
  as.data.frame() %>% 
  pull(label)

# Inspect
node_23335
```

```{r tree tibble for node 23109}
# Or if there are two nodes that we can see we can find its MRCA
## create tibble of tree to identify offspring
tree_tib <- as_tibble(unrooted_physeq@phy_tree)

# Inspect
head(tree_tib)

str(tree_tib)

# lets look at the long branch
mrca_node_23109 <- tidytree::MRCA(unrooted_tree, .node1 = 23109) 

# Inspect
mrca_node_23109 

## create dataframe to view ASVs and taxa info and to pull the ASVs
node_23109 <- 
  offspring(tree_tib, mrca_node_23109, tiponly = TRUE) %>% 
  as.data.frame() %>% 
  pull(label)

# Inspect
node_23109
```

```{r tree tibbl}
# # Or if there are two nodes that we can see we can find its MRCA
# ## create tibble of tree to identify offspring
# tree_tib <- as_tibble(unrooted_physeq@phy_tree)
# 
# # Inspect
# head(tree_tib)
# 
# str(tree_tib)
# 
# # lets look at the long branch
# mrca_node_23335 <- tidytree::MRCA(unrooted_tree, .node1 = 23335) 
# 
# # Inspect
# mrca_node_23335 
# 
# ## create dataframe to view ASVs and taxa info and to pull the ASVs
# node_23335 <- 
#   offspring(tree_tib, mrca_node_23335, tiponly = TRUE) %>% 
#   as.data.frame() %>% 
#   pull(label)
# 
# # Inspect
# node_23335
```

```{r pull taxonomy for 23335 and 23109}
# Now, let's pull out the taxonomy information, including the ASV sequence!
#### for 23335
node_23335_df <- 
  unrooted_physeq %>%
  tax_table() %>%
  as.data.frame() %>% 
  filter(ASV %in% node_23335) 

# Take a look: 
glimpse(node_23335_df)

#### for 23109
node_23109_df <- 
  unrooted_physeq %>%
  tax_table() %>%
  as.data.frame() %>% 
  filter(ASV %in% node_23109) 

# Take a look: 
glimpse(node_23109_df)
```


```{inspect ASV counts of 23335 and 23109}
# look at counts of the 23335 ASVs
unrooted_physeq %>%
  subset_taxa(., ASV %in% node_23335_df$ASV) %>%
  otu_table() %>%
  data.frame() %>%
  colSums()
# these were in 9 out of 60 samples, and the highest count was 241 in one sample, but the others are around 1-50 per sample


# look at counts of the 23109 ASVs
unrooted_physeq %>%
  subset_taxa(., ASV %in% node_23109_df$ASV) %>%
  otu_table() %>%
  data.frame() %>%
  colSums()

# these were in 4 out of 60 samples. One sample had 45 , but the others had 5-9 per sample. 
```


### Inspecting the one archaea 5251

```{r tree tibble for node 5251}
# Or if there are two nodes that we can see we can find its MRCA
## create tibble of tree to identify offspring
tree_tib <- as_tibble(unrooted_physeq@phy_tree)

# Inspect
head(tree_tib)

str(tree_tib)

# lets look at the long branch
mrca_node_5251 <- tidytree::MRCA(unrooted_tree, .node1 = 5251, ) 
ancestor(unrooted_tree, 5251) 

#create tibble to manipulate tree
node_test <- offspring(tree_tib, 23091, tiponly = TRUE) %>%
  as.data.frame()

# Inspect
mrca_node_5251 

## create dataframe to view ASVs and taxa info and to pull the ASVs
node_5251 <- 
  offspring(tree_tib, 23091, tiponly = TRUE) %>% 
  as.data.frame() %>% 
  pull(label)

# Filtering for only archaea, since we only have one archaea in this node
node_5251_df <- as.data.frame(unrooted_physeq@tax_table) %>% filter(ASV %in% node_5251) %>%
  filter(Kingdom == "Archaea")

node_5251_df %>% pull(ASVseqs)
```

I blased and I'm getting a lot of uncultured archaeon hits, which is good, and the BLAST E valu e is prettly low (-133) BUT the Query value is not high enough (65%). We want it to be 70% or higher. And this ASV clusters really high to bacteria while we would expect it to be more divergent. 
We also blasted stuff on the same branch as the closest ancestor, and they appeared to be uncultured bacteria with 100% query cover. So this further convinces me that the archaea is not a real data point and should be pruned.


```{r Pruning 23335 and 23109 and 5212}
# Function from Joey McMurdie: https://github.com/joey711/phyloseq/issues/652
pop_taxa = function(physeq, badTaxa){
  allTaxa <-  taxa_names(physeq)
  allTaxa <- allTaxa[!(allTaxa %in% badTaxa)]
  return(prune_taxa(allTaxa, physeq))}

# Let's use the pop_taxa function :) 
# Recreate a phyloseq object without your unwanted ASV or node
# Lets use the example ASVs above from node_3315_df
unrooted_physeq_rm_euk_arch <- 
  unrooted_physeq %>%
  pop_taxa(., node_23335_df$ASV) %>%
  pop_taxa(., node_23109_df$ASV) %>%
  pop_taxa(., node_5251_df$ASV)

# Check it 
unrooted_physeq_rm_euk_arch

# Intuition Check: How many differ???
ntaxa(unrooted_physeq) - ntaxa(unrooted_physeq_rm_euk_arch) # 17 nodes are different. Checks, since both nodes combined had 17 nodes (13 and 4 respectively)

# Visually inspect 
ggtree(unrooted_physeq_rm_euk_arch) + 
  geom_tippoint(mapping = aes(color = Kingdom)) +
  geom_text(aes(label=node), hjust= -0.5, vjust = -0.3, size = 2)


# the pruned tree is much better! It removed the Eukaryote and NA labels, which means we got all of them with the pruning. There are also no more too-long branches
```


INTERPRETATION #2: Were there any long branches in the unrooted tree? If so, did you decide to remove any ASVs from your tree? Why or why not? What data convinced you to remove each of the ASVs?

There were two long branches in the unrooted tree, full of eukaryotes and NA ASV's. The first branch was 23335, which had 13 non-bacteria or archaea ASVs, and the second was 23109, which had 4. After looking at the tibble and the number of samples containing ASV's in these nodes, I found that they were very rare, which raised my suspicions. Node 23335 was in 9 out of 60 samples, and the highest ASV count was 241 in one sample, but the others are around 1-50 per sample. Node 23109 was in 4 out of 60 samples. One sample had 45 counts, but the others had 5-9 per sample. 
Since they were rare, and also eukaryotic, I decided to prune them. Additionally, there was one archaea, but in the unrooted tree, it was a very short branch, and did not look diverged from other bacteria enough to be a real archaea. Even in the new tree after pruning the eukaryotic branches, the archaea branch was still very short. Therefore, I BLASTed it.
I got all uncultured archaeon hits, which was a positive sign, and additionallythe BLAST E-value is pretty low (-133). However, the Query value was not high enough (65%). We want it to be 70% or higher. Then, I checked the nearest other hits by BLASTing bacterial sequences on the same branch as the closest ancestor, and they appeared to be uncultured bacteria with 100% query cover. So this further convinced me that the archaea is not a real data point, and therefore I pruned it. 


# 5. Root the Tree

## 5b. Midpoint Root 

```{r midpoint-root-tree-physeq}
# Check to see if tree is rooted.. expect it to be false
is.rooted(phy_tree(unrooted_physeq_rm_euk_arch)) ##false

# Let's midpoint root the tree
midpoint_rooted_tree <- 
  unrooted_physeq_rm_euk_arch %>%
  phy_tree() %>%
  phytools::midpoint.root() 

# Is the new tree rooted? Should be TRUE!
is.rooted(midpoint_rooted_tree)

# Assign to a new phyloseq object!
# Merge tree with the raw_preprocessed physeq
midpoint_rooted_physeq <-
  merge_phyloseq(raw_preprocessed_physeq, midpoint_rooted_tree)
```


### Plot Midpoint Rooted Tree

```{r plot-midpoint-root-tree,fig.width=4.5, fig.height=7}
# Quick inspection of tree 
midpoint_rooted_tree_plot <-
  ggtree(midpoint_rooted_physeq) + 
  geom_tippoint(mapping = aes(color = Kingdom)) + 
  geom_tippoint(aes(color = Kingdom)) + 
  labs(title = "Midpoint Rooted Tree") + 
  scale_color_manual(values = c("cornflowerblue", "firebrick3", "grey"))

# Look at the tree 
midpoint_rooted_tree_plot

# Add nodes 
midpoint_rooted_tree_node_plot <- 
  midpoint_rooted_tree_plot + 
  geom_text(aes(label = node), hjust = -0.5, vjust = -0.3, size = 1.5) 

# View it 
midpoint_rooted_tree_node_plot
```

INTERPRETATION #3 (If midpoint rooting): Now that you’ve rooted your tree, does the tree seem “finalized”? Are there any other branches/ASVs to inspect? Why or why not?

I think the three seems finalized. There are no extremely long branches when rooted at the midpoint. The  tree is split into two big clusters, but even at the base, none of the branches are too long. There are no final branches to inspect.

# 7. Save phyloseq objects

The phyloseq objects can be saved in one file as a list of objects. Or, we can save them individually. Two examples are below: 

### Save multiple phyloseq objects as a list

This will keep two phyloseq objects in on .RData object that will allow us to easily load in both trees into an R session at once. 

```{r save-phyloseq-objects-list}
# Take a quick look
midpoint_rooted_physeq # Midpiont rooted tree

# Save both phyloseq objects to one .RData file 
save(list = c("midpoint_rooted_physeq"),
     file = "data/04_PhylogeneticTree/phytree_preprocessed_physeqs.RData")
```

### Save individual phyloseq objects

Here, we will create two individual files that will each have a single phyloseq object. This is useful if we only want to work with one tree. (For example, we will move forward with the archaeal tree.) 

```{r save-phyloseq-object}

# Now, save midpoint rooted phyloseq object
save(midpoint_rooted_physeq, 
     file = "data/04_PhylogeneticTree/midpoint_rooted_physeq.RData")
```

Now that we have our rooted trees inside of a phyloseq object, we can go forth and begin our next analysis!

# Final info for Reproducibility 

## Check Render Time
```{r stop-time}
# Take the time now that we are at the end of the script
end_time <- Sys.time()
end_time 

# Echo the elapsed time
elapsed_time <- round((end_time - start_time), 3)
elapsed_time
```


